---
title: "Before (and After) Roe v. Wade: A Content Analysis of Conservative and Liberal Media on Abortion Legislation in the United States, 2020–2024"
subtitle: "Exploring Sentiment, Framing, and State-Level Policy Intersections in U.S. News Coverage"
author: 
- "Erika Salvador"
- "Emilie Ward"
- "Maigan Lafontant"
abstract: "This blog investigates how abortion is framed in the media by analyzing coverage from two ideologically opposed outlets: Fox News and The New York Times. Using text and sentiment analysis, topic modeling, and geographic policy data from the Guttmacher Institute, we examine shifts in media narratives before and after the 2022 Supreme Court decision in Dobbs v. Jackson Women’s Health Organization."
editor: 
  markdown: 
    wrap: 72
bibliography: bib/library.bib
csl: bib/american-statistical-association.csl
---

```{r}
#| label: setup
#| include: false

# =================================================================
# Setup: Global Options and Libraries
# =================================================================

# Set code chunk defaults for cleaner output
knitr::opts_chunk$set(
  tidy = FALSE,                # Preserve code formatting
  size = "small",              # Use slightly smaller font for code
  message = FALSE,             # Suppress messages in output
  options(scipen = 1,          # Prefer standard numeric formatting (no scientific notation)
          knitr.kable.NA = '') # Display empty cells instead of "NA" in tables
)

# Load required libraries
library(tidytext)
library(dplyr)
library(widyr)
library(igraph)
library(ggraph)
library(tibble)
library(readr)
library(lubridate)  
library(ggridges)
library(gganimate)
library(magick)
library(ggplot2)
library(tidyr)
library(plotly)
library(forcats)
library(ggTimeSeries)
library(zoo)
library(data.table)
library(RColorBrewer)
library(scico)  
library(patchwork)
library(showtext)
library(ggiraph)
library(gdtools)
library(gfonts)

# Load cleaned datasets
nyt <- read_csv("../data/wrangled/nyt_news_data_wrangled.csv")
fox <- read_csv("../data/wrangled/fox_news_data_wrangled.csv")
## guttmacher <- read_csv("../data/Guttmatcher_institute_data/guttmacher_file.csv")  

# Combine NYT and Fox together
articles <- bind_rows(
  nyt |> mutate(outlet = "The New York Times"),
  fox |> mutate(outlet = "Fox News")
) |> 
  mutate(pub_date = as.Date(pub_date)) |> 
  filter(!is.na(pub_date))
```

```{r}
#| label: initial-tokens
#| include: false

# Tokenize text into individual words
tokens <- bind_rows(
  nyt |> mutate(outlet = "NYT"),
  fox |> mutate(outlet = "Fox")
) |>
  unnest_tokens(word, text) |>
  anti_join(stop_words)  # Remove standard English stop words first

# Additional stop words
custom_stopwords <- tibble(word = c(
  "said", "ms", "mr", "mrs",
  "fox", "news", "nyt", "new", "york", "times",
  "u.s", "one", "two", "also", "would", "including", "don’t", "that’s", "it’s", "wrote", "told",
  "roe"
))
```

```{r}
#| label: clean-tokens
#| include: false

# Tokenize and remove both standard and custom stop words
tokens2 <- bind_rows(
  nyt |> mutate(outlet = "NYT"),
  fox |> mutate(outlet = "Fox")
) |>
  unnest_tokens(word, text) |>
  anti_join(stop_words) |>         # Remove standard English stop words
  anti_join(custom_stopwords)      # Remove custom filler/irrelevant words
```

```{r}
#| label: monthly-data
#| include: false

monthly_data <- articles |> 
  mutate(month = floor_date(pub_date, unit = "month")) |> 
  count(outlet, month, name = "monthly_n") |> 
  arrange(outlet, month) |> 
  group_by(outlet) |> 
  mutate(cumulative_n = cumsum(monthly_n)) |> 
  ungroup() |>
  mutate(outlet = factor(outlet, levels = c("Fox News", "The New York Times")))


```

```{r}
#| label: color-choices-yay

# Color palette 
fill_colors <- c(
  "Fox News" = "#fccfd1",
  "The New York Times" = "#bfd7ea"
)

line_colors <- c(
  "Fox News" = "#e57373",
  "The New York Times" = "#5a8ca4"
)
```

```{r}
#| label: themes
#| include: false

# Load News Cycle from Google Fonts
font_add_google("News Cycle")
showtext_auto()

gdtools::register_gfont("News Cycle")
```

## Introduction

Long a topic of debate across law, medicine, gender, and religion, abortion remains one of the most polarizing issues in American public life. That polarization deepened in June 2022, when the Supreme Court’s decision in [*Dobbs v. Jackson Women’s Health Organization*](https://www.oyez.org/cases/2021/19-1392) overturned [*Roe v. Wade*](https://www.oyez.org/cases/1971/70-18), the 1973 ruling that had established a constitutional right to abortion. The decision unsettled not only the legal framework around abortion, but also the broader discourse surrounding it. As states move in sharply different directions, media coverage has had to grapple with new divisions, new language, and new stakes.

At its core, the public debate still turns on a simple but deeply charged question: **is abortion good or bad?** How that question gets answered—or even framed—depends heavily on where one looks. News coverage of abortion varies sharply by ideological leaning. Conservative media outlets often emphasize religious and legal arguments, focusing on fetal personhood, state-level bans, and moral appeals [@jenssen2013broadcast]. In contrast, liberal media frequently frame abortion in terms of bodily autonomy, healthcare access, and gender justice, while highlighting the disproportionate impact of abortion restrictions on marginalized communities [@rohlinger2015abortion].

In this blog, we examine how abortion is framed by two ideologically distinct media outlets: [Fox News](https://www.foxnews.com) and [The New York Times](https://www.nytimes.com). These platforms were selected for their national prominence, wide readership, and well-documented political leanings—Fox News is widely recognized as a conservative outlet, while The New York Times is identified as liberal-leaning [@pew2014; @mitchell2020].

We complement this analysis with geographic policy data from the [Guttmacher Institute](https://www.guttmacher.org), a leading research organization on reproductive health. Using tools such as sentiment analysis, topic modeling, and spatial comparison, we explore how media narratives align with state-level abortion laws and how discourse evolves in response to shifting legal landscapes.


## A First Look at the Coverage

We collected a total of `r nrow(articles)` abortion-related news
articles for this analysis. Of these, `r nrow(fox)` were published by
Fox News and the remaining `r nrow(nyt)` by The New York Times, spanning
from early 2020 through late 2024. Together, these articles provide a
window into how two major media outlets approached abortion coverage
across a period of significant legal and political change.

```{r}
#| label: eda 
#| warnings: false
#| mesages: false
#| include: false

monthly_data_2 <- monthly_data |>
  mutate(
    outlet = factor(outlet, levels = c("Fox News", "The New York Times")),
    
    # For monthly volume
    tooltip_text_monthly = paste0(outlet, "\nMonth: ", format(month, "%b %Y"), "\nArticles: ", monthly_n),
    data_id_monthly = paste(outlet, "Monthly", format(month, "%Y-%m"), monthly_n, sep = "_"),

    # For cumulative
    tooltip_text_c = paste0(outlet, "\nMonth: ", format(month, "%b %Y"), "\nArticles: ", cumulative_n),
    data_id_c = paste(outlet, "Cumulative", format(month, "%Y-%m"), cumulative_n, sep = "_")
  )

```

To trace how attention fluctuated over time, we first examined monthly trends in article publication. Both *Fox News* and *The New York Times* exhibited visible surges in abortion-related reporting across the 2020–2024 period. While our primary focus is on the [*Dobbs v. Jackson Women’s Health Organization*](https://www.supremecourt.gov/opinions/21pdf/19-1392_6j37.pdf) decision and its aftermath, earlier spikes in coverage are also notable. Attention intensified sharply in **September 2021** with the enactment of [Texas’s SB8 “heartbeat bill”](https://capitol.texas.gov/tlodocs/87R/billtext/pdf/SB00008I.pdf), surged again in **May 2022** when *Politico* [leaked the Supreme Court’s draft opinion](https://www.politico.com/news/2022/05/02/supreme-court-abortion-draft-opinion-00029473) in *Dobbs*, and peaked in **June 2022** when the Court [officially overturned *Roe v. Wade*](https://www.supremecourt.gov/opinions/21pdf/19-1392_6j37.pdf).

```{r}
#| label: monthly-plot
#| fig.height: 3.5

# Build the plot
monthly_plot <- ggplot(monthly_data_2, aes(x = month, y = monthly_n, group = outlet)) +
  geom_area_interactive(
    aes(fill = outlet, tooltip = tooltip_text_monthly, data_id = data_id_monthly),
    alpha = 0.35, position = "identity"
  ) +
  geom_line_interactive(
    aes(color = outlet, tooltip = tooltip_text_monthly, data_id = data_id_monthly),
    linewidth = 1.2
  ) +
  geom_vline(
    xintercept = as.Date("2022-06-24"),
    linetype = "dashed",
    color = "grey80",
    linewidth = 0.8
  ) +
  scale_fill_manual(values = fill_colors) +
  scale_color_manual(values = line_colors) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b'%y") +
  labs(
    title = "Monthly Volume of Abortion Coverage",
    x = "Date", y = "Number of Articles",
    fill = "Outlet", color = "Outlet"
  ) +
  theme_light(base_family = "News Cycle") +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(size = 6, angle = 90, hjust = 1),
    panel.border = element_rect(color = "grey80", fill = NA),
    plot.margin = margin(10, 15, 10, 15)
  )

# Render
girafe(
  ggobj = monthly_plot,
  options = list(
    opts_hover(css = "stroke-width:2;"),
    opts_tooltip(css = "background-color:white;padding:5px;border:1px solid gray;border-radius:4px;font-family:'News Cycle';"),
    opts_toolbar(saveaspng = TRUE),
    opts_sizing(rescale = TRUE)
  )
)
```

Beyond the immediate aftermath of *Dobbs*, two additional peaks in abortion-related media coverage align with major electoral moments. The first appears in **November 2022**, during the [U.S. midterm elections](https://www.npr.org/2022/11/09/1135345050/abortion-rights-state-ballots-election-results), when abortion rights were a central issue on state ballots across the country. Notably, *Fox News* overtook *The New York Times* in article volume — a reversal from earlier months when *The Times* had led during key judicial moments. This surge in Fox’s reporting likely reflects heightened partisan framing around abortion’s political salience, especially in contested states.

A second uptick occurs in **August 2023**, during the [Ohio special election](https://www.nytimes.com/2023/08/08/us/ohio-abortion-constitution-vote.html), when voters decided whether to raise the threshold for constitutional amendments. The measure was widely understood as a proxy battle over abortion rights. While both outlets covered the vote, the media response was more subdued than in 2022 — perhaps due to its more localized scope despite national implications.

The most recent spike emerges in **late 2024**, with *The New York Times* reclaiming the lead in coverage volume. This increase likely corresponds with renewed national focus on abortion in the lead-up to the [2024 U.S. presidential election](https://www.nytimes.com/interactive/2024/us/elections/calendar.html), where reproductive rights once again became a central campaign issue.

While monthly patterns highlight media reactions to key events, they don’t fully capture the broader arc of attention. To provide a cumulative perspective, we plotted the running total of abortion-related articles over time. Although *The New York Times* often displayed sharper monthly peaks, the cumulative view reveals a different pattern: over the long term, *Fox News* produced a significantly larger volume of abortion-related coverage. Following the *Dobbs* decision, both outlets showed a sharp rise in article counts, but *Fox News* continued publishing abortion-related content at a faster pace. By the end of 2024, *Fox News* had amassed a much larger body of coverage compared to *The New York Times*.

```{r}
#| label: cumulative-plot
#| fig.height: 3.5

# Build interactive plot
cumulative_plot <- ggplot(monthly_data_2, aes(x = month, y = cumulative_n, group = outlet)) +
  geom_area_interactive(
    aes(fill = outlet, tooltip = tooltip_text_c, data_id = data_id_c),
    alpha = 0.35, position = "identity"
  ) +
  geom_line_interactive(
    aes(color = outlet, tooltip = tooltip_text_c, data_id = data_id_c),
    linewidth = 1.2
  ) +
  geom_vline(
    xintercept = as.Date("2022-06-24"),
    linetype = "dashed",
    color = "grey80",
    linewidth = 0.8
  ) +
  scale_fill_manual(values = fill_colors) +
  scale_color_manual(values = line_colors) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b'%y") +
  labs(
    title = "Cumulative Abortion Coverage Over Time",
    x = "Date", y = "Cumulative Number of Articles",
    fill = "Outlet", color = "Outlet"
  ) +
  theme_light(base_family = "News Cycle") +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(size = 6, angle = 90, hjust = 1),
    panel.border = element_rect(color = "grey80", fill = NA),
    plot.margin = margin(10, 15, 10, 15)
  )

# Render
girafe(
  ggobj = cumulative_plot,
  options = list(
    opts_hover(css = "stroke-width:2;"),
    opts_tooltip(css = "background-color:white;padding:5px;border:1px solid gray;border-radius:4px;font-family:'News Cycle';"),
    opts_toolbar(saveaspng = TRUE),
    opts_sizing(rescale = TRUE)
  )
)
```

## A Calendar of Changing Narratives

To understand how language and sentiment around abortion shifted in the
media before and after the Supreme Court’s landmark decision in Dobbs v.
Jackson Women’s Health Organization, we mapped article sentiment scores
across the two major outlets in a calendar heatmap format spanning 2020
to 2024. Since we’ve been talking about how Fox News devoted a large volume of coverage to abortion, let’s first take a closer look at the tone of that coverage.

From 2020 to early 2024, Fox News’s sentiment trends tell a story of measured reporting occasionally pierced by sharp reactions. The early years — 2020 and 2021 — are relatively quiet. The sentiment tiles are mostly pale, which indicates mostly neutral-toned articles with only the occasional dip into negative or positive language. Abortion, at this point, seemed to be a non-headline driver, and Fox’s emotional engagement with the topic remained limited.

That changes dramatically in 2022. The calendar lights up in early summer, just as the Supreme Court decision in Dobbs v. Jackson Women’s Health Organization overturned Roe v. Wade. Around June, the heatmap becomes a patchwork of red and blue — days of intense negativity followed by equally charged positivity. The tonal whiplash reflects a polarized editorial moment: some coverage framed the ruling as a conservative victory; other stories emphasized the backlash, protests, and uncertainty that followed.

After the Dobbs ruling, the emotional peaks become more sporadic. In 2023 and 2024, the calendar shows fewer intense days, but when they appear, they’re sharper. It suggests that while abortion may not have remained a daily front-page story, the days it did appear were often tied to high-stakes developments — like new state laws, political campaigns, or anniversary reflections.

Unlike some coverage cycles, the sentiment doesn’t follow a weekly rhythm. Some Mondays are sharply negative, others are quiet. What we see instead is a reactive pattern: Fox’s sentiment spikes in response to external shocks, not internal scheduling. It’s not an even pulse; it’s a media heartbeat that jumps in moments of national tension.

So while the quantity of coverage tells us that Fox News paid close attention to abortion, the sentiment heatmap reveals when that attention turned emotional — and just how sharply it swung.

```{r}
#| label: fox-calendar-heatmap
#| fig.height: 6


# Tokenize and assign article ID
fox <- as.data.table(fox)

article_words <- fox |>
  select(pub_date, text_clean) |>
  mutate(article_id = row_number()) |>
  unnest_tokens(word, text_clean)

# Join with AFINN sentiment lexicon
afinn <- get_sentiments("afinn")

article_scores <- article_words |>
  inner_join(afinn, by = "word") |>
  group_by(article_id, pub_date) |>
  summarise(article_sentiment = sum(value), .groups = "drop")

# Floor timestamps, then average sentiment per day
fox_sentiment <- article_scores |>
  mutate(pub_day = as.Date(pub_date)) |>
  group_by(pub_day) |>
  summarise(sentiment_score = mean(article_sentiment), .groups = "drop") |>
  rename(date = pub_day) |>
  as.data.table()

# Create calendar covering all real article dates
calendar_fox <- data.table(
  date = seq(min(fox_sentiment$date), max(fox_sentiment$date), by = "day")
)

# Merge and interpolate missing days
calendar_fox <- merge(calendar_fox, fox_sentiment, by = "date", all.x = TRUE)
calendar_fox[, sentiment_score := na.approx(sentiment_score, na.rm = FALSE)]
calendar_fox[is.na(sentiment_score), sentiment_score := 0]

# Add calendar features
calendar_fox[, ':='(
  Year = as.integer(format(date, "%Y")),
  month = as.integer(format(date, "%m")),
  week = as.integer(format(date, "%V")),
  weekday = as.integer(format(date, "%u")),
  weekofyear = as.integer(format(date, "%V"))
)]
calendar_fox[is.na(week) | week <= 0, week := 1]
calendar_fox[is.na(weekofyear) | weekofyear <= 0, weekofyear := 1]

# Add interpretation
calendar_fox[, interpretation := fifelse(
  sentiment_score > 10, "Positive",
  fifelse(sentiment_score < -10, "Negative", "Neutral")
)]

# Create tooltip and data_id
calendar_fox[, tooltip := paste0(
  "Date: ", format(date, "%B %d, %Y"),
  "\nSentiment Score: ", round(sentiment_score, 2),
  "\nInterpretation: ", interpretation
)]
calendar_fox[, data_id := paste0("day_", date)]

# Dynamically compute sentiment range
range_max <- ceiling(max(abs(calendar_fox$sentiment_score), na.rm = TRUE))

# Build interactive ggplot
calendar_plot_fox <- ggplot(calendar_fox, aes(x = weekofyear, y = 8 - weekday)) +
  geom_tile_interactive(
    aes(
      fill = sentiment_score,
      tooltip = tooltip,
      data_id = data_id
    ),
    color = "grey80"
  ) +
  scale_fill_gradient2(
    low = "#cf1322",
    mid = "white",
    high = "#003f5c",
    midpoint = 0,
    limits = c(-range_max, range_max),
    breaks = c(-100, 0, 100),
    labels = c("Negative", "Neutral", "Positive"),
    name = "Sentiment"
  ) +
  facet_wrap(~Year, ncol = 1, strip.position = "right") +
  scale_y_continuous(
    name = NULL,
    breaks = 1:7,
    labels = c("Sun", "Sat", "Fri", "Thu", "Wed", "Tue", "Mon")
  ) +
  scale_x_continuous(
    name = "Week of Year",
    expand = c(0, 0)
  ) +
  theme_minimal(base_family = "News Cycle") +
  theme(
    legend.position = "bottom",
    axis.text = element_text(size = 8),
    strip.text = element_text(size = 13, face = "plain", color = "black"),
    plot.margin = margin(10, 15, 10, 15)
  )

# Render with girafe
girafe(
  ggobj = calendar_plot_fox,
  options = list(
    opts_hover(css = "fill-opacity:0.8;stroke:black;stroke-width:0.5px;"),
    opts_tooltip(css = "background-color:white;padding:5px;border:1px solid gray;border-radius:4px;font-family:'News Cycle';"),
    opts_toolbar(saveaspng = TRUE),
    opts_sizing(rescale = TRUE)
  )
)
```

After exploring how Fox News covered abortion in both volume and tone, it’s worth asking: how does The New York Times compare? While Fox’s sentiment calendar pulsed with emotional spikes tied to legal and political flashpoints, The Times paints a different picture.

Across five years of reporting, the Times’ calendar heatmap reveals a remarkably consistent tone, with most days hovering in the neutral-to-slightly-negative range. There’s none of the whiplash of bright reds or deep blues that characterized Fox’s reporting. Even in 2022 — the year of Dobbs v. Jackson — the Times maintains a calm visual register. A brief streak of dark red appears in late spring, which is likely tied to leaked drafts and early reactions to the Supreme Court’s decision, but it fades quickly. The overwhelming majority of tiles are beige or pale pink: stories that may be emotionally charged in content, but are tempered in language.

In fact, the New York Times appears to approach abortion coverage with a kind of editorial evenness. Unlike Fox News, which tends to react forcefully to political developments and court rulings, the Times reflects a more sustained and balanced journalistic rhythm. Coverage is spread consistently across the year, not just clustered around high-profile events, and sentiment rarely escalates into extremes.

What’s particularly striking is that one of the Times’ most positive sentiment spikes doesn’t occur after the Dobbs decision, but before it. On April 18, 2022, the sentiment score soars to 366.5 — unusually high, and notably upbeat for such a politically charged moment. One might expect coverage in this lead-up period to lean negative or cautious. Instead, this burst of positivity suggests a different editorial choice — perhaps highlighting stories of resilience, advocacy gains, or early signs of political mobilization.


```{r}
#| label: nyt-calendar-heatmap
#| warnings: false
#| messages: false
#| fig.height: 6

# Tokenize and assign article ID
nyt <- as.data.table(nyt)

article_words_nyt <- nyt |>
  select(pub_date, text_clean) |>
  mutate(article_id = row_number()) |>
  unnest_tokens(word, text_clean)

# Join with AFINN sentiment lexicon
afinn <- get_sentiments("afinn")

article_scores_nyt <- article_words_nyt |>
  inner_join(afinn, by = "word") |>
  group_by(article_id, pub_date) |>
  summarise(article_sentiment = sum(value), .groups = "drop")

# Floor timestamps, then average sentiment per day
nyt_sentiment <- article_scores_nyt |>
  mutate(pub_day = as.Date(pub_date)) |>
  group_by(pub_day) |>
  summarise(sentiment_score = mean(article_sentiment), .groups = "drop") |>
  rename(date = pub_day) |>
  as.data.table()

# Create calendar covering all real article dates
calendar_nyt <- data.table(
  date = seq(min(nyt_sentiment$date), max(nyt_sentiment$date), by = "day")
)

# Merge and interpolate missing days
calendar_nyt <- merge(calendar_nyt, nyt_sentiment, by = "date", all.x = TRUE)
calendar_nyt[, sentiment_score := na.approx(sentiment_score, na.rm = FALSE)]
calendar_nyt[is.na(sentiment_score), sentiment_score := 0]

# Add calendar features
calendar_nyt[, ':='(
  Year = as.integer(format(date, "%Y")),
  month = as.integer(format(date, "%m")),
  week = as.integer(format(date, "%V")),
  weekday = as.integer(format(date, "%u")),
  weekofyear = as.integer(format(date, "%V"))
)]
calendar_nyt[is.na(week) | week <= 0, week := 1]
calendar_nyt[is.na(weekofyear) | weekofyear <= 0, weekofyear := 1]

# Add interpretation
calendar_nyt[, interpretation := fifelse(
  sentiment_score > 10, "Positive",
  fifelse(sentiment_score < -10, "Negative", "Neutral")
)]

# Create tooltip and data_id
calendar_nyt[, tooltip := paste0(
  "Date: ", format(date, "%B %d, %Y"),
  "\nSentiment Score: ", round(sentiment_score, 2),
  "\nInterpretation: ", interpretation
)]
calendar_nyt[, data_id := paste0("day_", date)]

# Dynamically compute sentiment range
range_max_nyt <- ceiling(max(abs(calendar_nyt$sentiment_score), na.rm = TRUE))

# Build interactive ggplot
calendar_plot_nyt <- ggplot(calendar_nyt, aes(x = weekofyear, y = 8 - weekday)) +
  geom_tile_interactive(
    aes(
      fill = sentiment_score,
      tooltip = tooltip,
      data_id = data_id
    ),
    color = "grey80"
  ) +
  scale_fill_gradient2(
    low = "#cf1322",
    mid = "white",
    high = "#003f5c",
    midpoint = 0,
    limits = c(-range_max_nyt, range_max_nyt),
    breaks = c(-100, 0, 100),
    labels = c("More Negative", "Neutral", "More Positive"),
    name = "Sentiment"
  ) +
  facet_wrap(~Year, ncol = 1, strip.position = "right") +
  scale_y_continuous(
    name = NULL,
    breaks = 1:7,
    labels = c("Sun", "Sat", "Fri", "Thu", "Wed", "Tue", "Mon")
  ) +
  scale_x_continuous(
    name = "Week of Year",
    expand = c(0, 0)
  ) +
  theme_minimal(base_family = "News Cycle") +
  theme(
    legend.position = "bottom",
    axis.text = element_text(size = 8),
    strip.text = element_text(size = 13, face = "plain", color = "black"),
    plot.margin = margin(10, 15, 10, 15)
  )

# Render with girafe
girafe(
  ggobj = calendar_plot_nyt,
  options = list(
    opts_hover(css = "fill-opacity:0.8;stroke:black;stroke-width:0.5px;"),
    opts_tooltip(css = "background-color:white;padding:5px;border:1px solid gray;border-radius:4px;font-family:'News Cycle';"),
    opts_toolbar(saveaspng = TRUE),
    opts_sizing(rescale = TRUE)
  )
)
```



