---
title: "Before (and After) Roe v. Wade: A Content Analysis of Conservative and Liberal Media on Abortion Legislation in the United States, 2020–2024"
subtitle: "Exploring Sentiment, Framing, and State-Level Policy Intersections in U.S. News Coverage"
author: 
- "Erika Salvador"
- "Emilie Ward"
- "Maigan Lafontant"
abstract: "This blog investigates how abortion is framed in the media by analyzing coverage from two ideologically opposed outlets: Fox News and The New York Times. Using text and sentiment analysis, topic modeling, and geographic policy data from the Guttmacher Institute, we examine shifts in media narratives before and after the 2022 Supreme Court decision in Dobbs v. Jackson Women’s Health Organization."
editor: 
  markdown: 
    wrap: 72
bibliography: bib/library.bib
csl: bib/american-statistical-association.csl
---

```{r}
#| label: setup
#| include: false

# =================================================================
# Setup: Global Options and Libraries
# =================================================================

# Set code chunk defaults for cleaner output
knitr::opts_chunk$set(
  tidy = FALSE,                # Preserve code formatting
  size = "small",              # Use slightly smaller font for code
  message = FALSE,             # Suppress messages in output
  options(scipen = 1,          # Prefer standard numeric formatting (no scientific notation)
          knitr.kable.NA = '') # Display empty cells instead of "NA" in tables
)

# Load required libraries
library(tidytext)
library(dplyr)
library(widyr)
library(igraph)
library(ggraph)
library(tibble)
library(readr)
library(lubridate)  
library(ggridges)
library(gganimate)
library(magick)
library(ggplot2)
library(tidyr)
library(plotly)
library(forcats)
library(ggTimeSeries)
library(zoo)
library(data.table)
library(RColorBrewer)
library(scico)  
library(patchwork)
library(showtext)
library(ggiraph)
library(gdtools)
library(gfonts)
library(textmineR)
library(sf)
library(leaflet)
library(viridis)
library(ggspatial)
library(mapproj)
library(maps)
library(mapdata)
library(oz)
library(rnaturalearth)
library(urbnmapr)
library(ggplot2)
library(stringr)
library(fivethirtyeight)
library(plotly)

# Load cleaned datasets
nyt <- read_csv("../data/wrangled/nyt_news_data_wrangled.csv")
fox <- read_csv("../data/wrangled/fox_news_data_wrangled.csv")
## guttmacher <- read_csv("../data/Guttmatcher_institute_data/guttmacher_file.csv")  

# Combine NYT and Fox together
articles <- bind_rows(
  nyt |> mutate(outlet = "The New York Times"),
  fox |> mutate(outlet = "Fox News")
) |> 
  mutate(pub_date = as.Date(pub_date)) |> 
  filter(!is.na(pub_date))
```

```{r}
#| label: initial-tokens
#| include: false

# Tokenize text into individual words
tokens <- bind_rows(
  nyt |> mutate(outlet = "NYT"),
  fox |> mutate(outlet = "Fox")
) |>
  unnest_tokens(word, text) |>
  anti_join(stop_words)  # Remove standard English stop words first

# Additional stop words
custom_stopwords <- tibble(word = c(
  "said", "ms", "mr", "mrs",
  "fox", "news", "nyt", "new", "york", "times",
  "u.s", "one", "two", "also", "would", "including", "don’t", "that’s", "it’s", "wrote", "told",
  "roe"
))
```

```{r}
#| label: clean-tokens
#| include: false

# Tokenize and remove both standard and custom stop words
tokens2 <- bind_rows(
  nyt |> mutate(outlet = "NYT"),
  fox |> mutate(outlet = "Fox")
) |>
  unnest_tokens(word, text) |>
  anti_join(stop_words) |>         # Remove standard English stop words
  anti_join(custom_stopwords)      # Remove custom filler/irrelevant words
```

```{r}
#| label: monthly-data
#| include: false

monthly_data <- articles |> 
  mutate(month = floor_date(pub_date, unit = "month")) |> 
  count(outlet, month, name = "monthly_n") |> 
  arrange(outlet, month) |> 
  group_by(outlet) |> 
  mutate(cumulative_n = cumsum(monthly_n)) |> 
  ungroup() |>
  mutate(outlet = factor(outlet, levels = c("Fox News", "The New York Times")))


```

```{r}
#| label: color-choices-yay

# Color palette 
fill_colors <- c(
  "Fox News" = "#fccfd1",
  "The New York Times" = "#bfd7ea"
)

line_colors <- c(
  "Fox News" = "#e57373",
  "The New York Times" = "#5a8ca4"
)
```

```{r}
#| label: themes
#| include: false

# Load News Cycle from Google Fonts
font_add_google("News Cycle")
showtext_auto()

gdtools::register_gfont("News Cycle")
```

## Introduction

Long a topic of debate across law, medicine, gender, and religion,
abortion remains one of the most polarizing issues in American public
life. That polarization deepened in June 2022, when the Supreme Court’s
decision in [*Dobbs v. Jackson Women’s Health
Organization*](https://www.oyez.org/cases/2021/19-1392) overturned [*Roe
v. Wade*](https://www.oyez.org/cases/1971/70-18), the 1973 ruling that
had established a constitutional right to abortion. The decision
unsettled not only the legal framework around abortion, but also the
broader discourse surrounding it. As states move in sharply different
directions, media coverage has had to grapple with new divisions, new
language, and new stakes.

At its core, the public debate still turns on a simple but deeply
charged question: **is abortion good or bad?** How that question gets
answered—or even framed—depends heavily on where one looks. News
coverage of abortion varies sharply by ideological leaning. Conservative
media outlets often emphasize religious and legal arguments, focusing on
fetal personhood, state-level bans, and moral appeals
[@jenssen2013broadcast]. In contrast, liberal media frequently frame
abortion in terms of bodily autonomy, healthcare access, and gender
justice, while highlighting the disproportionate impact of abortion
restrictions on marginalized communities [@rohlinger2015abortion].

In this blog, we examine how abortion is framed by two ideologically
distinct media outlets: [Fox News](https://www.foxnews.com) and [The New
York Times](https://www.nytimes.com). These platforms were selected for
their national prominence, wide readership, and well-documented
political leanings—Fox News is widely recognized as a conservative
outlet, while The New York Times is identified as liberal-leaning
[@pew2014; @mitchell2020].

We complement this analysis with geographic policy data from the
[Guttmacher Institute](https://www.guttmacher.org), a leading research
organization on reproductive health. Using tools such as sentiment
analysis, topic modeling, and spatial comparison, we explore how media
narratives align with state-level abortion laws and how discourse
evolves in response to shifting legal landscapes.

## A First Look at the Coverage

We collected a total of `r nrow(articles)` abortion-related news
articles for this analysis. Of these, `r nrow(fox)` were published by
Fox News and the remaining `r nrow(nyt)` by The New York Times, spanning
from early 2020 through late 2024. Together, these articles provide a
window into how two major media outlets approached abortion coverage
across a period of significant legal and political change.

```{r}
#| label: eda 
#| warnings: false
#| mesages: false
#| include: false

# Create new dataset for plotting
monthly_data_2 <- monthly_data |>
  mutate(
    outlet = factor(outlet, levels = c("Fox News", "The New York Times")),

    # Tooltip & ID for monthly plot
    tooltip_text_monthly = paste0(
      outlet, "\nMonth: ", format(month, "%b %Y"),
      "\nArticles: ", monthly_n
    ),
    data_id_monthly = paste(outlet, "Monthly", format(month, "%Y-%m"), monthly_n, sep = "-"),

    # Tooltip & ID for cumulative plot
    tooltip_text_cumulative = paste0(
      outlet, "\nMonth: ", format(month, "%b %Y"),
      "\nArticles: ", cumulative_n
    ),
    data_id_cumulative = paste(outlet, "Cumulative", format(month, "%Y-%m"), cumulative_n, sep = "-")
  )

# Split for monthly plot
fox_monthly <- monthly_data_2 |> filter(outlet == "Fox News")
nyt_monthly <- monthly_data_2 |> filter(outlet == "The New York Times")

# Split for cumulative plot (reuse same df but different y-values)
fox_cumulative <- monthly_data_2 |> filter(outlet == "Fox News")
nyt_cumulative <- monthly_data_2 |> filter(outlet == "The New York Times")
```

To trace how attention fluctuated over time, we first examined monthly
trends in article publication. Both *Fox News* and *The New York Times*
exhibited visible surges in abortion-related reporting across the
2020–2024 period. While our primary focus is on the [*Dobbs v. Jackson
Women’s Health
Organization*](https://www.supremecourt.gov/opinions/21pdf/19-1392_6j37.pdf)
decision and its aftermath, earlier spikes in coverage are also notable.
Attention intensified sharply in **September 2021** with the enactment
of [Texas’s SB8 “heartbeat
bill”](https://capitol.texas.gov/tlodocs/87R/billtext/pdf/SB00008I.pdf),
surged again in **May 2022** when *Politico* [leaked the Supreme Court’s
draft
opinion](https://www.politico.com/news/2022/05/02/supreme-court-abortion-draft-opinion-00029473)
in *Dobbs*, and peaked in **June 2022** when the Court [officially
overturned *Roe v.
Wade*](https://www.supremecourt.gov/opinions/21pdf/19-1392_6j37.pdf).

```{r}
#| label: monthly-plot
#| fig.height: 3.5
#| warning: false
#| message: false 

monthly_plot <- ggplot() +
  geom_area_interactive(
    data = rbind(fox_monthly, nyt_monthly),
    aes(
      x = month, y = monthly_n,
      fill = outlet, tooltip = tooltip_text_monthly, data_id = data_id_monthly
    ),
    alpha = 0.35
  ) +
  geom_line_interactive(
    data = rbind(fox_monthly, nyt_monthly),
    aes(x = month, y = monthly_n, color = outlet),
    linewidth = 1.2
  ) +
  geom_point_interactive(
    data = rbind(fox_monthly, nyt_monthly),
    aes(x = month, y = monthly_n, color = outlet, tooltip = tooltip_text_monthly, data_id = data_id_monthly),
    size = 2, alpha = 0
  ) +
  geom_vline(
    xintercept = as.Date("2022-06-24"),
    linetype = "dashed", color = "grey80"
  ) +
  scale_fill_manual(values = fill_colors, drop = FALSE) +
  scale_color_manual(values = line_colors, drop = FALSE) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b'%y") +
  labs(
    title = "Monthly Volume of Abortion Coverage",
    x = "Date", y = "Number of Articles"
  ) +
  theme_light(base_family = "News Cycle") +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(size = 6, angle = 90, hjust = 1),
    panel.border = element_rect(color = "grey80", fill = NA),
    plot.margin = margin(10, 15, 10, 15)
  )

# Render
girafe(
  ggobj = monthly_plot,
  options = list(
    opts_hover(css = "stroke-width:2;"),
    opts_tooltip(css = "background:white;padding:5px;border:1px solid gray;border-radius:4px;"),
    opts_toolbar(saveaspng = TRUE)
  )
)
```

Beyond the immediate aftermath of *Dobbs*, two additional peaks in
abortion-related media coverage align with major electoral moments. The
first appears in **November 2022**, during the [U.S. midterm
elections](https://www.npr.org/2022/11/09/1135345050/abortion-rights-state-ballots-election-results),
when abortion rights were a central issue on state ballots across the
country. Notably, *Fox News* overtook *The New York Times* in article
volume — a reversal from earlier months when *The Times* had led during
key judicial moments. This surge in Fox’s reporting likely reflects
heightened partisan framing around abortion’s political salience,
especially in contested states.

A second uptick occurs in **August 2023**, during the [Ohio special
election](https://www.nytimes.com/2023/08/08/us/ohio-abortion-constitution-vote.html),
when voters decided whether to raise the threshold for constitutional
amendments. The measure was widely understood as a proxy battle over
abortion rights. While both outlets covered the vote, the media response
was more subdued than in 2022 — perhaps due to its more localized scope
despite national implications.

The most recent spike emerges in **late 2024**, with *The New York
Times* reclaiming the lead in coverage volume. This increase likely
corresponds with renewed national focus on abortion in the lead-up to
the [2024 U.S. presidential
election](https://www.nytimes.com/interactive/2024/us/elections/calendar.html),
where reproductive rights once again became a central campaign issue.

While monthly patterns highlight media reactions to key events, they
don’t fully capture the broader arc of attention. To provide a
cumulative perspective, we plotted the running total of abortion-related
articles over time. Although *The New York Times* often displayed
sharper monthly peaks, the cumulative view reveals a different pattern:
over the long term, *Fox News* produced a significantly larger volume of
abortion-related coverage. Following the *Dobbs* decision, both outlets
showed a sharp rise in article counts, but *Fox News* continued
publishing abortion-related content at a faster pace. By the end of
2024, *Fox News* had amassed a much larger body of coverage compared to
*The New York Times*.

```{r}
#| label: cumulative-plot
#| fig.height: 3.5
#| warning: false
#| message: false 

cumulative_plot <- ggplot() +
  geom_area_interactive(
    data = rbind(fox_cumulative, nyt_cumulative),
    aes(
      x = month, y = cumulative_n,
      fill = outlet, tooltip = tooltip_text_cumulative, data_id = data_id_cumulative
    ),
    alpha = 0.35
  ) +
  geom_line_interactive(
    data = rbind(fox_cumulative, nyt_cumulative),
    aes(x = month, y = cumulative_n, color = outlet),
    linewidth = 1.2
  ) +
  geom_point_interactive(
    data = rbind(fox_cumulative, nyt_cumulative),
    aes(x = month, y = cumulative_n, color = outlet, tooltip = tooltip_text_cumulative, data_id = data_id_cumulative),
    size = 2, alpha = 0
  ) +
  geom_vline(
    xintercept = as.Date("2022-06-24"),
    linetype = "dashed", color = "grey80"
  ) +
  scale_fill_manual(values = fill_colors, drop = FALSE) +
  scale_color_manual(values = line_colors, drop = FALSE) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b'%y") +
  labs(
    title = "Cumulative Abortion Coverage Over Time",
    x = "Date", y = "Cumulative Number of Articles"
  ) +
  theme_light(base_family = "News Cycle") +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.title = element_text(face = "bold", size = 18),
    axis.text.x = element_text(size = 6, angle = 90, hjust = 1),
    panel.border = element_rect(color = "grey80", fill = NA),
    plot.margin = margin(10, 15, 10, 15)
  )

# Render
girafe(
  ggobj = cumulative_plot,
  options = list(
    opts_hover(css = "stroke-width:2;"),
    opts_tooltip(css = "background:white;padding:5px;border:1px solid gray;border-radius:4px;"),
    opts_toolbar(saveaspng = TRUE)
  )
)

```

## A Calendar of Changing Narratives

To understand how language and sentiment around abortion shifted in the
media before and after the Supreme Court’s landmark decision in [*Dobbs
v. Jackson Women’s Health
Organization*](https://www.oyez.org/cases/2021/19-1392), we mapped
article [AFINN](https://en.wikipedia.org/wiki/AFINN) sentiment scores
across the two major outlets in a calendar heatmap format spanning 2020
to 2024.

Since we’ve been talking about how *Fox News* devoted a large volume of
coverage to abortion, let’s first take a closer look at the tone of that
coverage.

<iframe src="../echarts/interactive_calendar_fox.html" width="100%" height="690px" style="border: none;">

</iframe>

While *Fox News* consistently produced abortion-related articles
throughout the 2020–2024 period, its emotional tone varied more subtly
than expected. Most days register as light red or pale blue, which
indicates slightly negative or slightly positive sentiment, with fewer
extreme spikes than one might assume for such a polarizing topic.

In 2022, around the time of the *Dobbs* decision that overturned *Roe v.
Wade*, we observe scattered patches of deeper red and blue. However,
rather than clustering tightly around key events, these peaks appear
spread out across weeks.

> Interestingly, sentiment remained relatively stable in 2023 and 2024,
> with more diffuse tonal variation. This could imply that while the
> political salience of abortion persisted, the emotional framing of
> *Fox News* coverage settled into a more routine cadence post-*Dobbs*.

After exploring how *Fox News* covered abortion in both volume and tone,
it’s worth asking: **how does *The New York Times* compare?**

<iframe src="../echarts/interactive_calendar_NYT.html" width="100%" height="690px" style="border: none;">

</iframe>

Surprisingly, it’s *The New York Times* that reads as more restrained.
In contrast to *Fox’s* more visibly fluctuating sentiment, the *Times*
displays a remarkably steady emotional cadence in its abortion coverage
from 2020 through 2024. Most days are washed in muted reds and pale
blues. There is consistently neutral or mildly toned language across the
board.

The only noticeable deviation appears in spring 2022, just before the
Supreme Court officially decided *Dobbs v. Jackson Women’s Health
Organization*. Being a liberal-leaning outlet, and one generally
supportive of abortion rights, we expected the *New York Times* to lean
more negative in tone around the *Dobbs* decision. So it’s surprising to
see a spike in **positive sentiment** in the weeks *before* the ruling.

> That moment of optimism might feel out of place given the looming loss
> of constitutional protections — but it likely reflects the way the
> stories were framed: not with defeat, but with **mobilization**.
> Articles during that time may have highlighted state-level
> protections, legal challenges, or grassroots organizing, which could
> register as “positive” in sentiment scoring, even if the context was
> defensive or reactive.

For example, an article celebrating the swift passage of abortion
protections in certain states might score as "positive" based on
language even though it was prompted by a national-level loss.
Similarly, a story framing a restrictive abortion law as "historic" or
"a major success" (even if opposed editorially) could also register as
positive if the wording carried emotional approval.

Even as article volume rises again toward the end of 2024, likely
tracking with presidential election season, the emotional register stays
steady. The *New York Times* seems to maintain its editorial evenness
throughout.

# Running the Narrative Race

To visualize how *The New York Times’* framing of abortion evolved over
time, we animated the top bigrams appearing in abortion-related articles
each month from **2020** to **2024**.

<iframe src="../echarts/animated_bar_charts_nyt.html" width="100%" height="500px" style="border: none;">

</iframe>

In **early 2020**, bigrams like *health care*, *pro life*, and *women's
health* were prominent, reflecting a healthcare- and
ideologically-focused framing of abortion. In **September 2021**, with
[Texas’s SB8 law](https://en.wikipedia.org/wiki/Texas_Senate_Bill_8)
going into effect, terms like *texas law* and *planned parenthood*
surged.

By **May and June 2022**, during the [leaked draft opinion of *Dobbs v.
Jackson Women's Health
Organization*](https://www.politico.com/news/2022/05/02/supreme-court-abortion-draft-opinion-00029473)
and the eventual [Dobbs
decision](https://www.supremecourt.gov/opinions/21pdf/19-1392_6j37.pdf),
legal terms like *court decision* became prominent and terminology
surrounding *rights* and *laws* shaped the overall discourse, marking a
shift to more court-centric, legal framing of abortion discourse.

Following the **Dobbs** decision, phrases like *overturn roe* and *court
decision* remained dominant. At the same time, healthcare-focused
language persisted with *health care*, while *ballot measures* rose to
prominence during state-level campaigns in
[Kansas](https://www.kansascity.com/news/politics-government/article264005286.html)
and
[Ohio](https://www.npr.org/2023/11/08/1211521603/ohio-election-abortion-rights-issue-1).

The presence of terms like *12 weeks*, *15 weeks*, *shield laws*, and
*aid access* reflects the growing complexity of abortion policy
post-*Roe*, especially surrounding medication abortion and interstate
protections.

As the articles move through **2023** and **2024** mentions of *election
results*, *ballot measures* as well as the *republican sentate* and the
*democratic party* demonstrate the growingly politicized nature of
abortion articles. In **September 2024** as the election comes close
metions of *Kamala Harris*, *Donald Trump*, and *JD Vance* are at the
forefront of the discourse around abortion.

**Overall**, we see how *The New York Times* coverage adapted in real
time to political events—shifting from national healthcare debates to
increasingly legal and rights-based discourse.

We did the similar thing for Fox News.

<iframe src="../echarts/animated_bar_charts_fox.html" width="100%" height="500px" style="border: none;">

</iframe>

We consistently see *health care* as a primary focus each year, but
especially in **2020**, there was a heightened emphasis on *public
care*, with frequent mentions of phrases like *affordable care*, *care
access*, and *medical progress*.

A focus on the pregnancy and the fetus top bigrams, *unborn children*,
*fetal heartbeat*, and *partial birth* in in **February 2021**. Articles
in 2021 have an increasing focus on the pregnancy and the fetus when
disscusing abortion. While **2020** centered more on *medicine*, *public
health*, and social institutions impacted by abortion laws, **2021**
marks a shift toward individual experiences and situations of pregnancy,
mothers, and children.

During **May 2022** and \*\*June 2022\* the [*Dobbs v. Jackson Women's
Health Organization*
decision](https://www.supremecourt.gov/opinions/21pdf/19-1392_6j37.pdf)
is at the forefront of media with a large spike in *supreme court*,
*court decision*, and *court ruling* in the vocabulary of many articles.
After the descion there is an increased use in *anti/pro* discourse.
Terms such as pro life are matched with pro choice and anti abortion is
matched with pro abortion to reflect the focus on the polarized nature
of the abortion in the aftermath of the decision.

Throughout the entire four year, there are mentions of *communion*, and
the *catholic faith* in **November 2020**, *religious freedom*, in
**December 2020** and *satanic temple* in **Feburary 2023**, suggesting
relgious narrative and arguments surrounding abortion. We see more
striking terms such as *rape incest*, *gender identity*, and *15 weeks*
like we saw in the NYT articles, suggesting the complex narratives and
range of perspectives in the current media discussions about abortion.

This wider thematic spread reflects the surge in *media attention* and
*public discourse* in the lead-up to and aftermath of the overturning of
[*Roe v. Wade*](https://www.oyez.org/cases/1971/70-18)—capturing debates
around health care, law, religion, and personal trauma in equal measure.

# The Knots and Threads of Abortion Discourse

::: panel-tabset

### Fox – Pre-Roe

<iframe src="../echarts/network_fox_preroe.html" width="100%" height="500px" style="border: none;"></iframe>

### Fox – Post-Roe

<iframe src="../echarts/network_fox_postroe.html" width="100%" height="500px" style="border: none;"></iframe>

### NYT – Pre-Roe

<iframe src="../echarts/network_nyt_preroe.html" width="100%" height="500px" style="border: none;"></iframe>

### NYT – Post-Roe

<iframe src="../echarts/network_nyt_postroe.html" width="100%" height="500px" style="border: none;"></iframe>

:::


# Aditional Information About Abortion Legistation
After predominantly studying media, there in an overwhelming discussion about the supreme court case the the highly politicized nature of the legislation It is, while disusing the social and legal impacts of Roe v Wade is important, being aware of the legislation separate from narratives of mainstream media is equally as important. 

The [Guttmatcher Insitute](https://www.guttmacher.org/) helps to relay up to date information about legislation on abortion and abortion rights on a state wide level. After exploring 

::: panel-tabset

### Abortion Bans Map

```{r}
#| label: map plot 1
#| fig.height: 3.5
#| warning: false
#| message: false 

data_bans <- read.csv("/Users/emilieward/Documents/GitHub/blog-MEE/data/Guttmatcher_institute_data/abortion_bans_data.csv", header = TRUE)

state_map_coord <- maps::map("state", plot = FALSE, fill = TRUE) |>
  st_as_sf()

bans_full <- full_join(state_map_coord, data_bans, by = "ID")

bans_full <- bans_full %>%
  mutate(
    ID = str_to_title(ID),
    Gestational_Duration = str_to_title(Gestational_Duration),
    specifics = ifelse( specifics == "X", "", specifics)) |>
  mutate(specifics = if_else(specifics == "Viability", "Abortion Ban After Viability (22~24 weeks)", specifics))


pal <- colorFactor(
  palette = c("#C7E9C0", "#74C476", "#238B45", "#00441B", "grey80"),
  domain = bans_full$Gestational_Duration
)

leaflet(bans_full) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal(Gestational_Duration),
    color = "white",
    weight = 1,
    popup = ~paste(
      "<strong>State:</strong>", ID, "<br>",
      "<strong>Ban Type:</strong>", Gestational_Duration, "<br>",
      "<strong>Details:</strong>", specifics
    )
  )

```

### Abortion Exceptions Map
```{r, echo = FALSE}
#| label: map plot 2
#| fig.height: 3.5
#| warning: false
#| message: false 


data_exceptions <- read.csv("/Users/emilieward/Documents/GitHub/blog-MEE/data/Guttmatcher_institute_data/abortion_exceptions_data.csv", header = TRUE)

state_map_coord <- maps::map("state", plot = FALSE, fill = TRUE) |>
  st_as_sf()

exceptions_full <- left_join(state_map_coord, data_exceptions, by = "ID")

exceptions_full <- exceptions_full|>
  mutate(ID = str_to_title(ID)) |> 
  mutate(number_of_exceptions = if_else(is.na(number_of_exceptions), 0, number_of_exceptions))

pal <- colorFactor(
  palette = c("#C7E9C0", "#74C476", "#238B45", "#00441B", "grey80"),
  domain = exceptions_full$number_of_exceptions
)


leaflet(exceptions_full) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal(number_of_exceptions),
    color = "white",
    weight = 1,
    popup = ~paste(
      "<strong>State:</strong>", ID, "<br>",
      "<strong>Number of Exceptions:</strong>", number_of_exceptions, "<br>",
      "<strong>Exceptions:</strong><br>", exception_list
    )
  )

```


:::