---
title: "Wrangling NBC News Articles"
subtitle: "STAT 231: Blog App"
author: "Maigan Lafontant, Emilie Ward, and Erika Salvador"
format: 
  pdf
header-includes: |
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r}
#| label: setup
#| include: false

# set code chunk option defaults
knitr::opts_chunk$set(
  # display code as types
  tidy = FALSE, 
  # slightly smaller code font
  size = "small",
  # do not display messages in PDF
  message = FALSE,
# improve digit and NA display 
  options(scipen = 1, knitr.kable.NA = ''))

# Load required libraries
library(tidyverse)
library(httr)
library(readr)
library(stringr)
library(tibble)
library(kableExtra)
library(colorspace)
library(jsonlite)
library(lubridate)
```

## 1. Read and Prepare Raw JSON

The JSON file `fox_news_data.json` was scraped from Fox News using a Python-based Tor-enabled web scraper. It includes article-level information such as titles, dates, body text, links, and images.

```{r}
#| label: step-1

## =================================================================
# Step 1: Read and Prepare Raw JSON (flatten = FALSE)
# =================================================================
# Note: Setting flatten = FALSE is important here.
#       It preserves nested structure and prevents JSON explosion,
#       which allows for faster and more manageable parsing downstream.

# Load JSON scraped from Python
nbc_news <- fromJSON("../data/nbc_news_data.json", flatten = FALSE)

# Preview structure
glimpse(nbc_news)
```

## 2. Subset Metadata

In this step, we subset and clean the raw Fox News data to retain only the important metadata fields needed for our analysis. Specifically, we extract the article title, description, full text, publication date, URL, and author information. The `publicationDate` field is simplified by selecting only the first timestamp per article. Because the original `url` values are relative paths, we prepend the full domain (`https://www.nbcnews.com`) to make them complete. The `authors` column is nested and occasionally malformed, so we use safety checks to extract and collapse the authors' names into a single string or return `NA` if no valid data is found.

```{r}
#| label: step-2

# =================================================================
# Step 2: Subset Metadata
# =================================================================

nbc_meta <- nbc_news |>
  # Convert to tibble for easier manipulation
  as_tibble() |>
  transmute(
    title,  # Article headline

    authors = map_chr(authors, ~ {  # Safely extract author names
      if (is.null(.x) || !is.data.frame(.x) || ncol(.x) == 0) {
        NA_character_
      } else {
        paste(unlist(.x), collapse = ", ")
      }
    }),
    
    # Use the first timestamp
    publicationDate = map_chr(publicationDate, ~ .x[1]),  
    # Convert relative to absolute URLs
    url = paste0("https://www.nbcnews.com", url),         
    description,  # Short summary
    text          # Full article body
  )
```

## 3. Filter Articles by Publication Year (2020-2024)

In this step, we subset the dataset to include only articles published between 2020 and 2024. We first convert the `publicationDate` string into a proper datetime object using `lubridate::ymd_hms()`, then retain only those entries whose year falls within our desired range.

```{r}
#| label: step-3

```

## 4. Clean Article Text


```{r}
#| label: step-4

# =================================================================
# Step 4: Clean Article Text
# =================================================================


```

```
